---
title: "NASS Agriculture data across NEON AOP sites"
output:
  ioslides_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse); library(gridExtra)

pwd <- "/nfs/public-data/NEON_workshop_data/NEON"

acs <- read.csv(paste0(pwd, "/NEON_AOP_ACS.csv"), stringsAsFactors = FALSE)
nas <- read.csv(paste0(pwd,"/nassdatafinal.csv"), stringsAsFactors = FALSE)
```

```{r, include=FALSE}
lnas <- nas %>% 
        pivot_longer(-Site, names_to = "description", values_to = "amount")

fert <- lnas %>% 
        filter(description %in% c("Acres_Treated_Manure", "Acres_Treated_Organic_Fert", "Acres_Treated_Fert_Total"))
```

```{r barplots, include=FALSE}
b_plot <- function(df, y, z){
          pl <- ggplot(df, aes(x = Site, y = {{y}}, fill = {{z}})) +
                geom_col(position = "dodge")  + theme_bw() + 
                theme(axis.text.x = element_text(angle = 90, vjust = 0.5) ,
                      axis.text = element_text(size = 8),
                      axis.title = element_text(size = 16)#, 
          #           legend.text = element_text(size = 11)
                      )
          return(pl)
          }
```


# Crops and Fertilizer


## Corn Acres harvested
```{r}
corn <- b_plot(df = nas, y = Corn_Acres_Harvested, z = "gold") + 
        theme(legend.position = "none") + ylab("corn acres harvested") +
        scale_fill_manual(values = c("gold"))
corn
```

## Tomato Acres harvested
```{r}
toms <- b_plot(df = nas, y = Tomato_Acres_Harvested_County, z = "darkred") + 
        theme(legend.position = "none") +
        scale_fill_manual(values = c("darkred"))
toms
```

## Total acres treated with fertilizer
```{r}
tot_fert <- b_plot(df = nas, y = Acres_Treated_Fert_Total, z = "#1B9E77") + 
            theme(legend.position = "none") + ylab("total acres treated w/ fertilizer") +
            scale_fill_manual(values = c("#1B9E77"))
tot_fert
```

```{r, include=FALSE, eval=FALSE}
# print fertilizer figure
pdf("total_fertilizer_nass.pdf", width = 10, height = 6)
print(tot_fert)
dev.off()
```

## Acres treated with Organic fertilizer
```{r}
org_fert <- b_plot(df = nas, y = Acres_Treated_Organic_Fert, z = "#7570B3") + 
            theme(legend.position = "none") + 
            scale_fill_manual(values = c("#7570B3"))
org_fert
```

## Acres treated by fertilizer type
```{r}
all_fert <- b_plot(df = fert, y = amount, z = description) + 
            ylab("Acres") +
            scale_fill_brewer(palette = "Dark2") +
            theme(axis.text.x = element_text(angle = 90),
                  legend.title = element_blank(),
                  legend.position = c(0.9, 0.9)) 
all_fert
```

## Tomato and Fertilizer panel plot
```{r}
grid.arrange(toms, tot_fert)

```

## Corn and Fertilizer panel plot
```{r}
c_n <- grid.arrange(corn, tot_fert)
c_n
```

```{r, include=FALSE, eval=FALSE, warning=FALSE, message=FALSE}
# print fertilizer and corn figure
c_n_plot <- arrangeGrob(corn, tot_fert) #generates 
ggsave(file="corn_fertilizer_nass.pdf", c_n_plot) #saves 
```

## Crop and Total Fertilizer panel plot
```{r}
grid.arrange(corn, all_fert)
```


# Women-owned Agriculture


## Proportion of Ag land Woman-owned
```{r}
prop_w <- b_plot(df = nas, y = Percent_Ag_Land_In_Crop_Woman_Owned, z = "cyan4") + 
          theme(legend.position = "none") + ylab("proportion ag land woman owned") + 
          scale_fill_manual(values = c("cyan4"))
prop_w
```

```{r, include=FALSE, eval=FALSE}
# print woman-owned figure
pdf("prop_woman_owned_nass.pdf", width = 10, height = 6)
print(prop_w)
dev.off()
```

## Acres per Woman-owned Operation
```{r}
awo <- b_plot(df = nas, y = Acres_Per_WomanOwned_Operation, z = "cyan2") + 
          theme(legend.position = "none") + 
          scale_fill_manual(values = c("cyan2"))
awo
```

## Crop and Women-owned operations 
```{r}
grid.arrange(corn, awo)
```



# Income and Subsidies


## Gov't Dollars per Operation
```{r}
gd <- b_plot(df = nas, y = Govt_Dollars_Per_Operation, z = "darkgreen") + 
          theme(legend.position = "none") + 
          scale_fill_manual(values = c("darkgreen"))
gd
```


```{r, include=FALSE}
animal <- b_plot(df = nas, y = Number_Animal_Operations)
animal
```


```{r, include=FALSE}
grid.arrange(animal, gd)
```


## Net Income per Operation
```{r}
income <- b_plot(df = nas, y = Net_Income_Per_Operation, z = "lightgreen") + 
          theme(legend.position = "none") + ylab("net income per operation") +
          scale_fill_manual(values = c("lightgreen"))
income
```

```{r, include=FALSE, eval=FALSE}
# print net income figure
pdf("net_income_op_nass.pdf", width = 10, height = 6)
print(income)
dev.off()
```


## Crop and Net Income
```{r}
grid.arrange(toms, income)
```



